---
title: "nichevol: An R package for assessment of evolutionary change of ecological niches"
author: "Marlon E. Cobos, Hannah Owens, and A. Townsend Peterson"
output:
  github_document:
    toc: yes
    toc_depth: 4
---

```{r knitr_init, echo=FALSE, cache=FALSE, message=FALSE}
packages <- c("knitr", "raster", "rgdal")
suppressWarnings(s <- sapply(packages, function(x) {library(x, character.only = T)}))
opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

<br>

## Package description

The **nichevol** R package helps perform critical steps in the assessment of species' ecological niche evolution. The main analyses include: initial exploration of environmental data in species records and accessible areas, preparation of data for phylogenetic analyses, comparative phylogenetic analyses, and plotting for interpretations.

<br>

## Installing the package

The most recent version of **nichevol** is available as a GitHub repository and can be installed and/or loaded using the code below. 

Note: Try the code below first... If you have any problem during the installation, restart your R session, close other sessions you may have open, and try again. If during the installation you are asked to update packages, please do it. If any of the packages gives an error, please install it alone using install.packages(), then try re-installing **nichevol** again. Also, it may be a good idea to update R and RStudio (if you are using it).

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# Installing and loading packages
if(!require(devtools)){
    install.packages("devtools")
}
if(!require(nichevol)){
    devtools::install_github("marlonecobos/nichevol")
}
library(nichevol)
```

<br>

## Exploring the nichevol package

### Setting a directory

The main functions of **nichevol** produce results that need to be written in a directory in your computer. Writing the results outside the R environment helps to avoid problems related to RAM limitations. That is why setting a working directory is recommended before starting, as follows:

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
setwd("Drive:/Your/Directory") # change the characters accordingly
```

<br>

### Functions in nichevol

A complete list of the main functions in the **nichevol** package can be found in the package documentation. Use the following code to see the list.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
help(nichevol)
```
